Practical Effects of Integrating Temperature with Strang Split Reactions

### 2021-03-26 strang/paper.tex
address editor's comments

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
For astrophysical reacting flows, operator splitting is commonly used
to couple hydrodynamics and reactions.  Each process operates
independent of one another, but by staggering the updates in a
symmetric fashion (via Strang splitting) second order accuracy in time
can be achieved.  However, approximations are often made to the
reacting system, including the choice of whether or not to integrate
temperature with the species.  Here we demonstrate through a simple
convergence test that integrating an energy equation together with reactions
achieves the best convergence when modeling reactive flows with Strang
splitting.  Additionally, second order convergence cannot be achieved without integrating an energy or temperature equation.
\end{abstract}

@@

Simulations of stellar flows require solving the equations of
hydrodynamics coupled with a nuclear reaction network.  The equations
of hydrodynamics with reacting sources are:
\begin{align}
\frac{\partial \rho}{\partial t} + \nabla \cdot (\rho \Ub) &= 0 \\
\frac{\partial (\rho X_k)}{\partial t} + \nabla \cdot (\rho X_k \Ub) &= \rho \omegadot_k \\
\frac{\partial (\rho \Ub)}{\partial t} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &= 0 \\
\frac{\partial (\rho E)}{\partial t} + \nabla \cdot \left [ (\rho E + p) \Ub \right ] &= \rho \Sdot
\end{align}
where $\rho$ is the density, $\Ub$ is the velocity vector, $X_k$ are the species mass fractions
with creation rates $\omegadot_k$, $p$ is the pressure, $E$ is the
specific total energy, and $\Sdot$ is the nuclear energy generation rate.

@@

Figure~\ref{thefigure} shows the norm of the error vs.\ the coarse run
resolution.  The slope of these lines is a measure of the convergence
rate~\citep{ORVV}.  We see that all methods converge at least second order
for density, but for the thermodynamic quantities, $(\rho e)$ and $T$,
the method where only $X_k$ is evolved during reactions has larger
errors and much worse convergence than the other methods.  Looking at
the trace nuclei generated in the burning, $\isotm{C}{12}$ and
$\isotm{O}{16}$, we see a large difference between the two methods
that evolve some sort of energy and the one method where only $X_k$ is
evolved---the latter converging essentially first order at high
resolution.

>>>>

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
Many astrophysical environments involve convective or explosive flows
driven by thermonuclear reactions (Type Ia supernovae, classical
novae, X-ray bursts, stellar evolution).  Simulation codes need to
accurately capture the interactions between reactions and
hydrodynamics to produce realistic models of these events.  For
astrophysical reacting flows, operator splitting is commonly used to
couple hydrodynamics and reactions.  Each process operates independent
of one another, but by staggering the updates in a symmetric fashion
(via Strang splitting) second order accuracy in time can be achieved.
However, approximations are often made to the reacting system,
including the choice of whether or not to integrate temperature with
the species.  Here we demonstrate through a simple convergence test
that integrating an energy equation together with reactions achieves
the best convergence when modeling reactive flows with Strang
splitting.  Additionally, second order convergence cannot be achieved
without integrating an energy or temperature equation.
\end{abstract}

@@

Type Ia supernovae, classical novae, X-ray bursts, and general stellar
evolution reactive flows, where the nuclear energy release can drive
strong convective or explosive flows.  These stellar flows require solving the equations of
hydrodynamics coupled with a nuclear reaction network.  The equations
of hydrodynamics with reacting sources are:
\begin{align}
\frac{\partial \rho}{\partial t} + \nabla \cdot (\rho \Ub) &= 0 \\
\frac{\partial (\rho X_k)}{\partial t} + \nabla \cdot (\rho X_k \Ub) &= \rho \omegadot_k \\
\frac{\partial (\rho \Ub)}{\partial t} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &= 0 \\
\frac{\partial (\rho E)}{\partial t} + \nabla \cdot \left [ (\rho E + p) \Ub \right ] &= \rho \Sdot
\end{align}
where $\rho$ is the density, $\Ub$ is the velocity vector, $X_k$ are the species mass fractions
with creation rates $\omegadot_k$, $p$ is the pressure, $E$ is the
specific total energy, and $\Sdot$ is the nuclear energy generation rate.

@@

Figure~\ref{thefigure} shows the norm of the error vs.\ the coarse run
resolution.  The slope of these lines is a measure of the convergence
rate~\citep{ORVV}.  We see that all methods converge to at least second order
for density, but for the thermodynamic quantities, $(\rho e)$ and $T$,
the method where only $X_k$ is evolved during reactions has larger
errors and much worse convergence than the other methods.  Looking at
the trace nuclei generated in the burning, $\isotm{C}{12}$ and
$\isotm{O}{16}$, we see a large difference between the two methods
that evolve some sort of energy and the one method where only $X_k$ is
evolved---the latter converging essentially first order at high
resolution.
